A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-10 shanghai.tex
Removed the marginal comments on issues that have been addressed. Added compositio files to the repository.

Now we localize $C(G)$.
Let us say that $(g,T)$ in $C(G)$ is a \emph{weak isomorphism} if $g$ is surjective
and $T$ is an isomorphism of vector spaces.  Recall from
\cite{kashiwara-schapira:CatgoriesSheaves}*{Def 7.1.5} that a \emph{left multiplicative system}
is a family of morphisms $W$ that
\begin{enumerate}
\item contains all isomorphisms,
\item is closed under composition,
\item is closed under pullback by arbitrary morphisms in $C(G)$,
\item and if the cokernel \David{Is $C(G)$ abelian?} of a morphism $\phi \in \mor C(G)$ is in
$W$ then so is the kernel of $\phi$.
\end{enumerate}

@@

\begin{lemma}
The family of weak isomorphisms in $C(G)$ is a saturated, left multiplicative system.
\end{lemma}
\begin{proof}
Any isomorphism $(g, T)$ in $C(G)$ has both $g$ and $T$ isomorphisms, and thus is a weak isomorphism.
Clearly, the composition of two weak isomorphisms is a weak isomorphism.  The remaining properties
follow from the fact that in a weak isomorphism $(g, T)$, $g$ is an \'etale cover, together with standard
facts on \'etale covers and pullbacks. \David{Is this sufficient?}
\end{proof}
Let $W$ be the family of weak isomorphisms and \cdef{$C(G)[W^{-1}]$} be the localization of $C(G)$ at $W$.
A morphism in $C(G)[W^{-1}]$ is an equivalence class of diagrams
%  \begin{equation}\label{eq:morGCf}
\[   
\begin{tikzcd}[column sep=30]
   (f,\psi) \arrow{r}{(g,T)} & (f',\psi') & \arrow[swap]{l}{(h,U)} (f_1,\psi_1)
   \end{tikzcd}
\]
%  \end{equation}
 where $(g,T) : (f,\psi) \to (f',\psi')$ is any map in $C(G)$ 
 and $(h,U) :  (f',\psi') \to (f_1,\psi_1)$ is a weak isomorphism in $C(G)$ \cite{kashiwara-schapira:CatgoriesSheaves}*{Thm 7.1.16}.
%Likewise, $C(G)[W^{-1}]$ is a tensor category.

@@

To prove the next proposition, we will need a certain perspective on local systems.
 If $\qcs{E}$ is any local system on a connected scheme $G$ and $\bg$ is a geometric point on $G$,
 we may associate to $\qcs{E}$ a character of the \'etale fundamental group $\pi_1(G, \bg)$ \David{add citation}.
 Since $\pi_1(G, \bg)$ is defined as a projective limit of automorphism groups of finite \'etale covers, a general
 local system will correspond to a projective limit of characters of such groups.  But some local systems can be
 described in terms of a single cover $f: H \to G$ and a representation $\psi: \ker f \to \Aut(V)$;
 two such descriptions yield isomorphic local systems if and only if they are related by a weak
 isomorphism in $C(G)$.  While the description of local systems in terms of $\pi_1(G, \bg)$ does
 not generalize well to non-connected $G$, the description in terms of $C(G)$ does.  \David{Look okay?}

@@

 For any map $(g,T)$ in $C(G)$, let $L(g,T) : L(f,\psi) \to L(f',\psi')$ be the homomorphism of local systems defined 
 by $s \mapsto s' \ceq T\circ s\circ g$. 
 This construction defines a full monoidal functor $L$ from $C(G)$ to the category $\Loc(G)$ of local systems on $G$ that
 commutes with pullback along any morphism of $\QC$-schemes,
 as defined using Lemma~\ref{lemma:finite-pullback}.
 %
 \Clifton{There's detail missing here; I might want to include some of it.}
 %
 Moreover, since $L(f,T)$ is an isomorphism of local systems
 if and only if $(f,T)$ is a weak isomorphism, 
 the full functor $L : C(G) \to \Loc(G)$ factors through the localization functor $C(G) \to C(G)[W^{-1}]$ 
 to define a full and faithful functor $L_{W} : C(G)[W^{-1}] \to \Loc(G)$;
 see \cite{kashiwara-schapira:CatgoriesSheaves}*{Exercise 7.5}.

@@

  We now show that $[\tau_\qcs{L}]$ depends only on the isomorphism class of $\qcs{L}$.
  Let $\alpha : \qcs{L} \to \qcs{L'}$ be an isomorphism and
  let $\tau$ and $\tau'$ be any cocycles representing the classes in the
  image of $\qcs{L}$ and $\qcs{L'}$ under \eqref{cohomologous}; by construction,
  $\tau$ and $\tau'$ are defined by the choice of sections $s$ and $s'$.
  Now, $\varphi_x(s(x))$ is a
  non-zero scalar multiple of $s'(x)$; define $a: G(\bFq) \to \EEx$
  by $\alpha_x(s(x)) = a(x) s'(x)$.
  By the second part of \ref{qc.4},
  $a(x+y) = a(x)a(y)$, so $a \in G(\bFq)^*$. By the first part of
  \ref{qc.4}, 
  $\tau'(\Frob{}) = \tau(\Frob{}) \cdot (\,^{\Frob{}^{-1}}a \cdot a^{-1})$,
  %\Clifton{Fix this.}
  so $\tau$ and $\tau'$ are cohomologous. This shows that
  \begin{equation}\label{pich1}
  \begin{split}
    \QCiso{G} &\to \Hh^1(\Weil{},G(\bFq)^*),\\
    [\qcs{L}] &\mapsto [\tau_\qcs{L}]
  \end{split}
  \end{equation}
  is well-defined.

@@

\begin{lemma} \label{lem:dual-inv}
Let $X$ be an abelian group equipped with an action of $\Weil{}$, and write $\Frob{} = \Frob{}$.  Then
\begin{align*}
 (X^*)_{\Frob{}} &\to (X^{\Frob{}})^* \\
 [f] &\mapsto f|_{X^{\Frob{}}}
\end{align*}
is an isomorphism.
\end{lemma}
\begin{proof}
We can describe $X^{\Frob{}}$ as the kernel of the map $X \xrightarrow{\Frob{}-1} X$;
let $Y = (\Frob{}-1)X$ be the augmentation ideal.  Dualizing the sequence
\[
 0 \to X^{\Frob{}} \to X \to Y \to 0
\]
yields
\[
 0 \to Y^* \to X^* \to (X^{\Frob{}})^* \to \Ext^1_\ZZ(Y, \EEx).
\]
Since $\Ext^1_\ZZ(-,\EEx)$ vanishes, we get a natural isomorphism from the cokernel of $Y^* \xrightarrow{\Frob{}-1} X^*$ to $(X^{\Frob{}})^*$.
\end{proof}

@@

The pullback $\pi_0^*(\qcs{E})$ of $\qcs{E}$ along $\pi_0 : G \to \pi_0(G)$ is constant 
on geometric connected components, with stalks given by 
$(\pi_0^* \qcs{E})_g = \qcs{E}_{x}$ for all $g\in \bG^x$.  Thus both $\pi_0^*\qcs{E}$ and $\qcs{L}$
are constant on geometric connected components of $G$. 
The choices above define isomorphisms
$\gqcs{L}\vert_{\bG^x} \to  (\gqcs{E}_{x})_{\bG^x}$ for each $x\in \pi_0(\bG)$.
The resulting isomorphism $\gqcs{L} \to \pi_0^* \gqcs{E}$ satisfies \ref{QC.4}, 
thus defining an isomorphism $\qcs{L} \to \pi_0^* \qcs{E}$.
\end{proof}

@@

\begin{proof}
We may define full and faithful functors $F_n : \QC(\GN{T}_n) \to \QC(\GN{T})$
for each $n$ by $F_n(\qcs{L}) \ceq [\qcs{L}_n]$ and $F_n(\alpha_n) \ceq [\alpha_n]$.
For positive integers $n \le m$ these functors fit together into a diagram
\[
\begin{tikzcd}
\ & & & \QC(\GN{T})\\
\arrow{urrr}{F_n} \QC(\GN{T}_n) \arrow[swap]{rr}{f_{n\leq m}^*} && \QC(\GN{T}_m) \arrow[swap]{ur}{F_m}. & 
\end{tikzcd}
\]
Passing to isomorphism classes of objects and invoking the 
universal property of $\varinjlim_{n\in \ZZ_{>0}} \QCiso{\GN{T}_n}$ 
gives a group isomorphism
$\theta: \varinjlim_{n\in \ZZ_{>0}} \QCiso{\GN{T}_n} \to \QCiso{\GN{T}}$
with inverse $[\qcs{L}]_{/\text{iso}} \mapsto [\qcs{L}_{/\text{iso}}]$.
Now use Theorem~\ref{thm:application}, Proposition~\ref{prop:pullback} 
and the fact that  $\Hom_{\text{qc}}(T(K),\EEx) = \varinjlim_{n\in \ZZ_{>0}} \GN{T}_n(k)^*$ 
with reference to the family of group homomorphisms 
dual to $f_{n} : \GN{T}_{n+1}(\Fq) \to \GN{T}_n(\Fq)$.
\end{proof}

>>>>

Now we localize $C(G)$.
Let us say that $(g,T)$ in $C(G)$ is a \emph{weak isomorphism} if $g$ is surjective
and $T$ is an isomorphism of vector spaces.  Recall from
\cite{kashiwara-schapira:CatgoriesSheaves}*{Def 7.1.5} that a \emph{left multiplicative system}
is a family of morphisms $W$ that
\begin{enumerate}
\item contains all isomorphisms,
\item is closed under composition,
\item is closed under pullback by arbitrary morphisms in $C(G)$,
\item and if the cokernel \David{Is $C(G)$ abelian?} \Clifton{Yes. In fact, that's stated, implicitly, in the paragraph above: a tensor category is an abelian category.} of a morphism $\phi \in \mor C(G)$ is in
$W$ then so is the kernel of $\phi$.
\end{enumerate}

@@

\begin{lemma}
The family of weak isomorphisms in $C(G)$ is a saturated, left multiplicative system.
\end{lemma}
\begin{proof}
Any isomorphism $(g, T)$ in $C(G)$ has both $g$ and $T$ isomorphisms, and thus is a weak isomorphism.
Clearly, the composition of two weak isomorphisms is a weak isomorphism.  The remaining properties
follow from the fact that in a weak isomorphism $(g, T)$, $g$ is an \'etale cover, together with standard
facts on \'etale covers and pullbacks. \David{Is this sufficient?}\Clifton{Yes, I think so.}
\end{proof}
Let $W$ be the family of weak isomorphisms and \cdef{$C(G)[W^{-1}]$} be the localization of $C(G)$ at $W$.
A morphism in $C(G)[W^{-1}]$ is an equivalence class of diagrams
%  \begin{equation}\label{eq:morGCf}
\[   
\begin{tikzcd}[column sep=30]
   (f,\psi) \arrow{r}{(g,T)} & (f',\psi') & \arrow[swap]{l}{(h,U)} (f_1,\psi_1)
   \end{tikzcd}
\]
%  \end{equation}
 where $(g,T) : (f,\psi) \to (f',\psi')$ is any map in $C(G)$ 
 and $(h,U) :  (f',\psi') \to (f_1,\psi_1)$ is a weak isomorphism in $C(G)$ \cite{kashiwara-schapira:CatgoriesSheaves}*{Thm 7.1.16}.
%Likewise, $C(G)[W^{-1}]$ is a tensor category.

@@

To prove the next proposition, we will need a certain perspective on local systems.
 If $\qcs{E}$ is any local system on a connected scheme $G$ and $\bg$ is a geometric point on $G$,
 we may associate to $\qcs{E}$ a character of the \'etale fundamental group $\pi_1(G, \bg)$ \David{add citation}.\Clifton{I'll find one.}
 Since $\pi_1(G, \bg)$ is defined as a projective limit of automorphism groups of finite \'etale covers, a general
 local system will correspond to a projective limit of characters of such groups.  But some local systems can be
 described in terms of a single cover $f: H \to G$ and a representation $\psi: \ker f \to \Aut(V)$;
 two such descriptions yield isomorphic local systems if and only if they are related by a weak
 isomorphism in $C(G)$.  While the description of local systems in terms of $\pi_1(G, \bg)$ does
 not generalize well to non-connected $G$, the description in terms of $C(G)$ does.  \David{Look okay?}

@@

 For any map $(g,T)$ in $C(G)$, let $L(g,T) : L(f,\psi) \to L(f',\psi')$ be the homomorphism of local systems defined 
 by $s \mapsto s' \ceq T\circ s\circ g$. 
 This construction defines a full monoidal functor $L$ from $C(G)$ to the category $\Loc(G)$ of local systems on $G$ that
 commutes with pullback along any morphism of $\QC$-schemes,
 as defined using Lemma~\ref{lemma:finite-pullback}.
 Moreover, since $L(f,T)$ is an isomorphism of local systems
 if and only if $(f,T)$ is a weak isomorphism, 
 the full functor $L : C(G) \to \Loc(G)$ factors through the localization functor $C(G) \to C(G)[W^{-1}]$ 
 to define a full and faithful functor $L_{W} : C(G)[W^{-1}] \to \Loc(G)$;
 see \cite{kashiwara-schapira:CatgoriesSheaves}*{Exercise 7.5}.

@@

  We now show that $[\tau_\qcs{L}]$ depends only on the isomorphism class of $\qcs{L}$.
  Let $\alpha : \qcs{L} \to \qcs{L'}$ be an isomorphism and
  let $\tau$ and $\tau'$ be any cocycles representing the classes in the
  image of $\qcs{L}$ and $\qcs{L'}$ under \eqref{cohomologous}; by construction,
  $\tau$ and $\tau'$ are defined by the choice of sections $s$ and $s'$.
  Now, $\varphi_x(s(x))$ is a
  non-zero scalar multiple of $s'(x)$; define $a: G(\bFq) \to \EEx$
  by $\alpha_x(s(x)) = a(x) s'(x)$.
  By the second part of \ref{qc.4},
  $a(x+y) = a(x)a(y)$, so $a \in G(\bFq)^*$. By the first part of
  \ref{qc.4}, 
  $\tau'(\Frob{}) = \tau(\Frob{}) \cdot (\,^{\Frob{}^{-1}}a \cdot a^{-1})$,
  so $\tau$ and $\tau'$ are cohomologous. This shows that
  \begin{equation}\label{pich1}
  \begin{split}
    \QCiso{G} &\to \Hh^1(\Weil{},G(\bFq)^*),\\
    [\qcs{L}] &\mapsto [\tau_\qcs{L}]
  \end{split}
  \end{equation}
  is well-defined.

@@

\begin{lemma} \label{lem:dual-inv}
Let $X$ be an abelian group equipped with an action of $\Weil{}$.
% and write $\Frob{} = \Frob{}$. 
 Then
\begin{align*}
 (X^*)_{\Frob{}} &\to (X^{\Frob{}})^* \\
 [f] &\mapsto f|_{X^{\Frob{}}}
\end{align*}
is an isomorphism.
\end{lemma}
\begin{proof}
We can describe $X^{\Frob{}}$ as the kernel of the map $X \xrightarrow{\Frob{}-1} X$;
let $Y = (\Frob{}-1)X$ be the augmentation ideal.  Dualizing the sequence
\[
 0 \to X^{\Frob{}} \to X \to Y \to 0
\]
yields
\[
 0 \to Y^* \to X^* \to (X^{\Frob{}})^* \to \Ext^1_\ZZ(Y, \EEx).
\]
Since $\Ext^1_\ZZ(-,\EEx)$ vanishes, we get a natural isomorphism from the cokernel of $Y^* \xrightarrow{\Frob{}-1} X^*$ to $(X^{\Frob{}})^*$.
\end{proof}

@@

The pullback $\pi_0^*(\qcs{E})$ of $\qcs{E}$ along $\pi_0 : G \to \pi_0(G)$ is constant 
on geometric connected components, with stalks given by 
$(\pi_0^* \qcs{E})_g = \qcs{E}_{x}$ for all $g\in \bG^x$.  Thus both $\pi_0^*\qcs{E}$ and $\qcs{L}$
are constant on geometric connected components of $G$. 
The choices above define isomorphisms
$\gqcs{L}\vert_{\bG^x} \to  (\gqcs{E}_{x})_{\bG^x}$ for each $x\in \pi_0(\bG)$.
The resulting isomorphism $\gqcs{L} \to \pi_0^* \gqcs{E}$ satisfies \ref{QC.4}, 
thus defining an isomorphism $\qcs{L} \to \pi_0^* \qcs{E}$ in $\QC(G)$.
\end{proof}

@@

\begin{proof}
We may define full and faithful functors $F_n : \QC(\GN{T}_n) \to \QC(\GN{T})$
for each $n$ by $F_n(\qcs{L}) \ceq [\qcs{L}_n]$ and $F_n(\alpha_n) \ceq [\alpha_n]$.
For positive integers $n \le m$ these functors fit together into a diagram
\[
\begin{tikzcd}
\ & & & \QC(\GN{T})\\
\arrow{urrr}{F_n} \QC(\GN{T}_n) \arrow[swap]{rr}{f_{n\leq m}^*} && \QC(\GN{T}_m) \arrow[swap]{ur}{F_m}. & 
\end{tikzcd}
\]
Passing to isomorphism classes of objects and invoking the 
universal property of direct limits gives a group isomorphism
$\theta: \varinjlim_{n\in \ZZ_{>0}} \QCiso{\GN{T}_n} \to \QCiso{\GN{T}}$
with inverse $[\qcs{L}]_{/\text{iso}} \mapsto [\qcs{L}_{/\text{iso}}]$.
Now use Theorem~\ref{thm:application}, Proposition~\ref{prop:pullback} 
and the fact that  $\Hom_{\text{qc}}(T(K),\EEx) = \varinjlim_{n\in \ZZ_{>0}} \GN{T}_n(k)^*$, where this last direct limit it taken
with reference to the family of group homomorphisms 
dual to $f_{n} : \GN{T}_{n+1}(\Fq) \to \GN{T}_n(\Fq)$.
\end{proof}
